---
title: "RMarkdown demo"
author: "Rudolf Schlechter"
date: "2024-05-16"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

library(tidyverse)
library(broom)

```

##    First things first
###   Why do we care about R Markdown?

Scripts can be messy and difficult to follow when not enough information is provided. However, *too much* text in a script can be distracting!

R Markdown is a simple formatting syntax. The main purpose is to connect data analysis with documentation. For that, the most simple way is to combine **markdown** (md) with **R** syntax.

`.Rmd file` -> `dynamic document` -> `reproducible research` 

## Markdown

#     Header 1
##    Header 2
###   Header 3
####  Header 4
##### Header ...

Plain text.

You can use *italics* and **bold**.

superscript^s^ and subscript~s~.

~~strikethrough~~

- Unordered list
- Itemised
  - Item 1
  - Item 2

1. Ordered list

`verbatim code`

> block quote

equation: $e^{i/pi} + 1 = 0$

Table

| Component | Amount | Unit |
|:---------:|:------:|:----:|
| Buffer | 10 | $\mu$L |
| Primer Fwd | 1.5 | $\mu$L |
| Primer Fwd | 1.5 | $\mu$L |
| Water | 6 | $\mu$L |
| DNA | 1.0 | $\mu$L |

## Workflow
1. Open new .Rmd file
2. Embed code
3. Write text
4. Set output format (header)
5. Save and render
6. Share!

## Let's work with a dataset

To insert a code chunk, try `alt + ctrl + I` or `option + cmd + I`.

Let's install the [penguins data set](https://allisonhorst.github.io/palmerpenguins/)
```{r}
install.packages("palmerpenguins")
```

Now we load the package
```{r}
library(palmerpenguins)
data(package = 'palmerpenguins')
```

### Data summary
```{r}
str(penguins)
```

Looking at the structure of the data, we have three species of penguins: `r levels(penguins$species)`, found in the islands of `r levels(penguins$island)`.

```{r}
head(penguins)
```

The dataset contains data of `r nrow(penguins)` penguins, and each entry has phenotypic data such as  bill length (mm), bill depth (mm), flipper length (mm), body mass (g), sex (male/female), and year of entry.

```{r}
penguins %>% 
  na.exclude %>% 
  group_by(species, island, sex) %>% 
  summarise(bill_length = mean(bill_length_mm),
            bill_depth = mean(bill_depth_mm),
            flipper_length = mean(flipper_length_mm),
            body_mass = mean(body_mass_g))
```

We can make a nicer table using `kable` from the `knitr` package. Also we can get rid of the message with local options (set `message = FALSE`)

```{r, message=FALSE}
penguins %>% 
  na.exclude %>% 
  group_by(species, island, sex) %>% 
  summarise(bill_length = mean(bill_length_mm),
            bill_depth = mean(bill_depth_mm),
            flipper_length = mean(flipper_length_mm),
            body_mass = mean(body_mass_g)) %>% 
  knitr::kable()
```

## Exploratory analysis

R Markdown is useful to add those extra steps that might not end up in your final manuscript, such as your exploratory analysis!

Let's evaluate if the data is normally distributed by looking at histograms

```{r}
penguins %>% 
  pivot_longer(cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)) %>% 
  ggplot(aes(x = value)) +
  facet_wrap(~name, scales = "free_x")+
  geom_histogram()
```

We can analyse a bit further the body mass of the different species of penguins.

##    Linear model

A linear model is a regression analysis in which a response variable can be described as a function of one or more predictors. In R, we express it as `y ~ x`.

In our case, we will use body mass as our response variable and we want to see how the different species contribute to the variation in weight among penguins.

```{r}
# Model definition
mod1 = lm(body_mass_g ~ species, data = penguins)

# Model Summary
summary(mod1)

# Residuals
plot(resid(mod1))

# Extract some the data from the model
df.mod1 <- tidy(mod1) %>% 
  mutate(species = levels(penguins$species),
         mean_body_mass = ifelse(term == "(Intercept)", 
                                 estimate, 
                                 estimate[species == "Adelie"] + estimate))

```

We can extract information from our model and use inline code. For example, we have now learned that the weight of species `r df.mod1$species[1]` is lower (mean = `r df.mod1$mean_body_mass[1]` g) than that of species `r df.mod1$species[3]` (mean = `r df.mod1$mean_body_mass[3]` g), with a *p*-value = `r df.mod1$p.value[3]` and an adjusted *R*^2^ = `r summary(mod1)$adj.r.squared * 100`%.

Let's plot our results:

```{r}
penguins %>% 
  na.exclude %>% 
  ggplot(aes(x = species, y = body_mass_g))+
  geom_jitter(width = 0.1)+
  theme_bw()+
  labs(x = "Species", y = "Body mass [g]")
  
```

We can include the means extracted from the model:

```{r}
penguins %>% 
  na.exclude %>% 
  ggplot(aes(x = species, y = body_mass_g))+
  geom_jitter(width = 0.1)+
  # Add a red dot representing the mean of each group
  geom_point(data = df.mod1, aes(x = species, y = mean_body_mass),
             color = "red",
             size = 3)+
  theme_bw()+
  labs(x = "Species", y = "Body mass [g]")
```


